---
title: "Seattle Mariners 2026"
subtitle: "Stats, refreshed the morning after every game"
date: today
page-layout: full
code-fold: true
---

```{r setup, include=FALSE}
source(here::here("R", "00_setup.R"))

# In CI the data is fetched once before the render. When this page is rendered
# standalone (e.g. in Positron) the .rds files may not exist yet — fetch then.
if (!file.exists(here("data", "batting_2026.rds"))) {
  source(here::here("R", "01_fetch_data.R"))
}

batting  <- readRDS(here("data", "batting_2026.rds"))
pitching <- readRDS(here("data", "pitching_2026.rds"))

refreshed <- format(Sys.time(), tz = "America/Los_Angeles",
                    format = "%B %e, %Y at %l:%M %p %Z")

# Note when the underlying numbers came from the synthetic fallback rather than
# the live feed, so the page is never silently wrong.
src <- tryCatch(readLines(here("data", "data_source.txt"), n = 1),
                error = function(e) "unknown")
```

::: {.refreshed-badge}
Last refreshed: `r refreshed`
:::

```{r live-warning, echo=FALSE, results='asis'}
if (!identical(src, "live")) {
  cat("> ⚠️ **Heads up:** these numbers are from the synthetic fallback — the live",
      "FanGraphs / MLB Stats API feed was unavailable at render time.\n")
}
```

This site treats a baseball season like a bioinformatics experiment: players are
*samples*, their offensive stats are *features*, and we look for the hitters and
pitchers behaving differently from the rest. The numbers come from FanGraphs and
the MLB Stats API and are re-rendered automatically the morning after each
Mariners game.

[**Read the full differential analysis →**](analysis.qmd)

## Season at a glance

The biggest gaps between what hitters have actually produced (**wOBA**) and what
their batted-ball quality says they *should* have produced (**xwOBA**). Positive =
out-performing the profile (regression candidates down); negative = under-performing
(due for positive regression).

```{r glance}
glance <- batting |>
  dplyr::transmute(
    Player = player,
    Pos    = pos,
    PA     = pa,
    wOBA   = round(woba, 3),
    xwOBA  = round(xwoba, 3),
    Diff   = round(woba - xwoba, 3)
  ) |>
  dplyr::arrange(dplyr::desc(Diff))

over  <- utils::head(glance, 5)
under <- glance |> utils::tail(5) |> dplyr::arrange(Diff)
```

::: {layout-ncol="2"}

::: {}
**Running hot — over-performing**

```{r over}
knitr::kable(over, row.names = FALSE)
```
:::

::: {}
**Running cold — under-performing**

```{r under}
knitr::kable(under, row.names = FALSE)
```
:::

:::

## Pitching: who's been lucky?

ERA minus xERA — negative means a pitcher has allowed fewer runs than his
peripherals predict.

```{r pitching-glance}
pitching |>
  dplyr::transmute(
    Player = player,
    Role   = role,
    IP     = ip,
    ERA    = round(era, 2),
    xERA   = round(xera, 2),
    `ERA − xERA` = round(era_minus_xera, 2)
  ) |>
  dplyr::arrange(`ERA − xERA`) |>
  knitr::kable(row.names = FALSE)
```

---

*How this updates: a scheduled GitHub Actions workflow checks each morning whether
the Mariners finished a game the night before. If so, it re-fetches the stats,
re-renders this site, and deploys it to Vercel. Off-days are skipped, and the
synthetic fallback is never published.*
