---
title: "Seattle Mariners 2026"
subtitle: "Stats, refreshed the morning after every game"
date: today
page-layout: full
code-fold: true
---

```{r setup, include=FALSE}
source(here::here("R", "00_setup.R"))
source(here::here("R", "charts.R"))

# In CI the data is fetched once before the render. When this page is rendered
# standalone the .rds files may not exist yet — fetch then.
if (!file.exists(here("data", "batting_2026.rds"))) {
  source(here::here("R", "01_fetch_data.R"))
}

batting  <- readRDS(here("data", "batting_2026.rds"))
pitching <- readRDS(here("data", "pitching_2026.rds"))

refreshed <- format(Sys.time(), tz = "America/Los_Angeles",
                    format = "%B %e, %Y at %l:%M %p %Z")

# Note when the underlying numbers came from the synthetic fallback rather than
# the live feed, so the page is never silently wrong.
src <- tryCatch(readLines(here("data", "data_source.txt"), n = 1),
                error = function(e) "unknown")
```

::: {.refreshed-badge}
Last refreshed: `r refreshed`
:::

```{r live-warning, echo=FALSE, results='asis'}
if (!identical(src, "live")) {
  cat("> ⚠️ **Heads up:** these numbers are from the synthetic fallback — the live",
      "FanGraphs / MLB Stats API feed was unavailable at render time.\n")
}
```

This site treats a baseball season like a bioinformatics experiment: players are
*samples*, their offensive stats are *features*, and we look for the hitters and
pitchers behaving differently from the rest. The numbers come from FanGraphs and
the MLB Stats API and are re-rendered automatically the morning after each
Mariners game.

[**Read the full differential analysis →**](/analysis)

## Who's hot, who's not

**Hard contact vs results.** Does hitting the ball hard translate into production?
The dashed line is the league-wide trend; the grey band is its 95% confidence
interval (how sure we are of that *average* relationship — not a per-player normal
range). Players well *above* the line are out-producing their contact quality
(lucky or crafty); well *below*, under-producing (unlucky or poor launch angles).

```{r hard-hit, fig.width = 9, fig.height = 6}
chart_hard_hit(batting)
```

**BABIP — luck on balls in play.** League average is ~.300; sustained extremes are
rare, so red bars tend to cool off and navy bars tend to warm up.

```{r babip, fig.width = 9, fig.height = 5}
chart_babip(batting)
```

**Pitching: ERA vs the peripherals.** Negative bars mean a pitcher has allowed
fewer runs than his strikeouts, walks, and contact quality would predict.

```{r pitching-luck, fig.width = 9, fig.height = 5}
chart_pitching_luck(pitching)
```

## The numbers

The biggest gaps between what hitters have actually produced (**wOBA**) and what
their batted-ball quality says they *should* have produced (**xwOBA**).

```{r glance}
glance <- batting |>
  dplyr::transmute(
    Player = player,
    Pos    = pos,
    PA     = pa,
    wOBA   = round(woba, 3),
    xwOBA  = round(xwoba, 3),
    Diff   = round(woba - xwoba, 3)
  ) |>
  dplyr::arrange(dplyr::desc(Diff))

over  <- dplyr::slice_max(glance, Diff, n = 5, with_ties = FALSE)
under <- dplyr::slice_min(glance, Diff, n = 5, with_ties = FALSE) |>
  dplyr::arrange(Diff)
```

::: {layout-ncol="2"}

::: {}
**Running hot — over-performing**

```{r over}
knitr::kable(over, row.names = FALSE)
```
:::

::: {}
**Running cold — under-performing**

```{r under}
knitr::kable(under, row.names = FALSE)
```
:::

:::

**Pitching staff — ERA − xERA**

```{r pitching-glance}
pitching |>
  dplyr::transmute(
    Player = player,
    Role   = role,
    IP     = ip,
    ERA    = round(era, 2),
    xERA   = round(xera, 2),
    `ERA − xERA` = round(era_minus_xera, 2)
  ) |>
  dplyr::arrange(`ERA − xERA`) |>
  knitr::kable(row.names = FALSE)
```

---

*How this updates: a scheduled GitHub Actions workflow checks each morning whether
the Mariners finished a game the night before. If so, it re-fetches the stats,
re-renders this site, and deploys it to Vercel. Off-days are skipped, and the
synthetic fallback is never published.*
