---
title: "Seattle Mariners 2026"
subtitle: "Stats, refreshed the morning after every game"
date: today
page-layout: full
code-fold: true
---

```{r setup, include=FALSE}
source(here::here("R", "00_setup.R"))
source(here::here("R", "charts.R"))

# In CI the data is fetched once before the render. When this page is rendered
# standalone the .rds files may not exist yet — fetch then.
if (!file.exists(here("data", "batting_2026.rds"))) {
  source(here::here("R", "01_fetch_data.R"))
}

batting  <- readRDS(here("data", "batting_2026.rds"))
pitching <- readRDS(here("data", "pitching_2026.rds"))

refreshed <- format(Sys.time(), tz = "America/Los_Angeles",
                    format = "%B %e, %Y at %l:%M %p %Z")

# Note when the underlying numbers came from the synthetic fallback rather than
# the live feed, so the page is never silently wrong.
src <- tryCatch(readLines(here("data", "data_source.txt"), n = 1),
                error = function(e) "unknown")
```

::: {.refreshed-badge}
Last refreshed: `r refreshed`
:::

```{r live-warning, echo=FALSE, results='asis'}
if (!identical(src, "live")) {
  cat("> ⚠️ **Heads up:** these numbers are from the synthetic fallback — the live",
      "FanGraphs / MLB Stats API feed was unavailable at render time.\n")
}
```

This site treats a baseball season like a bioinformatics experiment: players are
*samples*, their offensive stats are *features*, and we look for the hitters and
pitchers behaving differently from the rest. The numbers come from FanGraphs and
the MLB Stats API and are re-rendered automatically the morning after each
Mariners game.

[**Read the full differential analysis →**](/analysis)

## Stat glossary

New to the sabermetric alphabet soup? Here's what every stat on this page means.

::: {.stat-glossary}
<dl>
<dt>wOBA — weighted on-base average</dt>
<dd>One number that captures a hitter's total offensive value. Instead of treating
every time on base equally (like OBP does), it weights each outcome — walk, single,
double, triple, home run — by how much that event is actually worth in run-scoring
terms. It's scaled to look like OBP, so league average is about **.320**, a good
hitter is around **.370**, and an elite one pushes **.400+**. Higher is better.</dd>

<dt>xwOBA — expected weighted on-base average</dt>
<dd>What a hitter <em>should</em> have produced based on <strong>how</strong> they hit
the ball, not whether it found a glove. Statcast feeds the exit velocity and launch
angle of every batted ball (plus sprint speed on grounders) into a model that returns
the wOBA a typical hitter earns on contact like that. Because it strips out luck,
defense, and ballpark, comparing it to actual wOBA is revealing: <strong>wOBA − xwOBA
above zero</strong> means a hitter is out-producing their contact quality (riding hot,
likely to cool off), while <strong>below zero</strong> flags strong contact that hasn't
paid off yet (an upside regression candidate).</dd>

<dt>BABIP — batting average on balls in play</dt>
<dd>Batting average counting only balls put in play — home runs and strikeouts are
excluded. League average sits near <strong>.300</strong>, and sustained extremes are
rare, so a sky-high BABIP usually signals good luck or favorable sequencing that
regresses, and a very low one signals the opposite.</dd>

<dt>ERA / xERA — earned run average vs. expected</dt>
<dd><strong>ERA</strong> is the earned runs a pitcher allows per nine innings — the
classic bottom-line number. <strong>xERA</strong> is the contact-quality-based
expectation, derived the same way as xwOBA. <strong>ERA − xERA</strong> separates skill
from luck: a negative gap means a pitcher has allowed fewer runs than their stuff
predicts (possibly lucky), a positive gap the reverse.</dd>

<dt>Hard-hit %, exit velocity & launch angle — the batted-ball inputs</dt>
<dd><strong>Exit velocity</strong> is how fast the ball leaves the bat (mph);
<strong>hard-hit %</strong> is the share of batted balls hit at 95 mph or more — a proxy
for raw contact quality. <strong>Launch angle</strong> is the vertical angle off the
bat: too low is a grounder, too high is a pop-up, and the productive line-drive window
sits in between. Together these are the ingredients xwOBA is built from.</dd>

<dt>PA / IP — sample size</dt>
<dd><strong>Plate appearances</strong> (hitters) and <strong>innings pitched</strong>
(pitchers). They aren't performance stats — they tell you how much to trust the rest.
Early in the season small samples swing wildly, so a gaudy line on 30 PA means far
less than the same line on 300.</dd>
</dl>
:::

## Who's hot, who's not

**Hard contact vs results.** Does hitting the ball hard translate into production?
The dashed line is the league-wide trend; the grey band is its 95% confidence
interval (how sure we are of that *average* relationship — not a per-player normal
range). Players well *above* the line are out-producing their contact quality
(lucky or crafty); well *below*, under-producing (unlucky or poor launch angles).

```{r hard-hit, fig.width = 9, fig.height = 6}
chart_hard_hit(batting)
```

**BABIP — luck on balls in play.** League average is ~.300; sustained extremes are
rare, so red bars tend to cool off and navy bars tend to warm up.

```{r babip, fig.width = 9, fig.height = 5}
chart_babip(batting)
```

**Pitching: ERA vs the peripherals.** Negative bars mean a pitcher has allowed
fewer runs than his strikeouts, walks, and contact quality would predict.

```{r pitching-luck, fig.width = 9, fig.height = 5}
chart_pitching_luck(pitching)
```

## The numbers

The biggest gaps between what hitters have actually produced (**wOBA**) and what
their batted-ball quality says they *should* have produced (**xwOBA**).

```{r glance}
glance <- batting |>
  dplyr::transmute(
    Player = player,
    Pos    = pos,
    PA     = pa,
    wOBA   = round(woba, 3),
    xwOBA  = round(xwoba, 3),
    Diff   = round(woba - xwoba, 3)
  ) |>
  dplyr::arrange(dplyr::desc(Diff))

over  <- dplyr::slice_max(glance, Diff, n = 5, with_ties = FALSE)
under <- dplyr::slice_min(glance, Diff, n = 5, with_ties = FALSE) |>
  dplyr::arrange(Diff)
```

::: {layout-ncol="2"}

::: {}
**Running hot — over-performing**

```{r over}
knitr::kable(over, row.names = FALSE)
```
:::

::: {}
**Running cold — under-performing**

```{r under}
knitr::kable(under, row.names = FALSE)
```
:::

:::

**Pitching staff — ERA − xERA**

```{r pitching-glance}
pitching |>
  dplyr::transmute(
    Player = player,
    Role   = role,
    IP     = ip,
    ERA    = round(era, 2),
    xERA   = round(xera, 2),
    `ERA − xERA` = round(era_minus_xera, 2)
  ) |>
  dplyr::arrange(`ERA − xERA`) |>
  knitr::kable(row.names = FALSE)
```

## Player pages on Baseball Savant

Want the pitch-by-pitch Statcast detail behind these numbers? Each player links to
their [Baseball Savant](https://baseballsavant.mlb.com) page.

```{r savant-helper, include=FALSE}
# Emit a markdown bullet list of Savant links for a data frame that has
# `player` + `mlbam_id`. Players without an id are skipped.
savant_bullets <- function(df, kind) {
  if (!"mlbam_id" %in% names(df)) return(invisible())
  urls <- savant_url(df$player, df$mlbam_id, kind)
  ok   <- !is.na(urls)
  cat(paste0("- [", df$player[ok], "](", urls[ok], ")"), sep = "\n")
  cat("\n")
}
```

::: {layout-ncol="2"}

::: {.savant-list}
### Hitters

```{r savant-hitters, results='asis'}
batting |>
  dplyr::arrange(dplyr::desc(pa)) |>
  savant_bullets("hitting")
```
:::

::: {.savant-list}
### Pitchers

```{r savant-pitchers, results='asis'}
pitching |>
  dplyr::arrange(dplyr::desc(ip)) |>
  savant_bullets("pitching")
```
:::

:::

---

*How this updates: a scheduled GitHub Actions workflow checks each morning whether
the Mariners finished a game the night before. If so, it re-fetches the stats,
re-renders this site, and deploys it to Vercel. Off-days are skipped, and the
synthetic fallback is never published.*
